Finiteness in Cubical Type Theory

### 2020-06-30 sections/finiteness-predicates.tex
finished prose for section 2.1

One of the nice things about containers is it gives us a generic way to define
``membership'':
\begin{multicols}{2}
  \begin{agdalisting} \label{container-membership}
    \ExecuteMetaData[agda/Container/Membership.tex]{membership-def}
  \end{agdalisting} \columnbreak
  \begin{agdalisting}
    \ExecuteMetaData[agda/Snippets/Introduction.tex]{fiber}
  \end{agdalisting}
\end{multicols} \noindent 
Here we're using the homotopy-theory notion of a \AgdaDatatype{fiber} to define
membership: a fiber for some function \(f\) and some point \(y\) in its codomain
is a value \(x\) and a proof that \(f \; x \;\AgdaDatatype{≡}\;y\).
Membership also makes more sense when described concretely in terms of lists:
\(x\;\AgdaDatatype{\ensuremath{\in}}\;\mathit{xs}\) means ``there is an index
into \(\mathit{xs}\) such that the index points at an item equal to \(x\)''.
\subsection{Split Surjections}
Now that we have our terms defined, let's look a little at how split
enumerability relates to more traditional, classical notions of finiteness.
In a classical setting we likely wouldn't mention ``lists'' or the like, and
would instead define finiteness based on the existence of some injection or
surjection, say a surjection from a finite prefix of the natural numbers.
In HoTT, surjections (or, more precisely, \emph{split} surjections
\cite[definition 4.6.1]{hottbook}), are defined like so:
\begin{multicols}{2}
  \begin{agdalisting} \label{sp-surj-eqn}
    \ExecuteMetaData[agda/Function/Surjective/Base.tex]{split-surjective}
  \end{agdalisting} \columnbreak
  \begin{agdalisting} \label{sp-surj-arrow-eqn}
    \ExecuteMetaData[agda/Function/Surjective/Base.tex]{split-surjection}
  \end{agdalisting}
\end{multicols}
As it turns out, our definition of finiteness here is precisely the same as a
surjection-based one, in quite a deep way!
\begin{lemma} \label{split-enum-is-split-surj}%
  A proof of split enumerability is equivalent to a split surjection from a
  finite prefix of the natural numbers.
  \begin{agdalisting}
    \ExecuteMetaData[agda/Cardinality/Finite/SplitEnumerable.tex]{is-split-inj-type}
  \end{agdalisting}
\end{lemma}
\begin{proof} \let\qed\relax \vspace{-1\baselineskip}
  \begin{minipage}[t]{.7\textwidth}\vspace{-\baselineskip}
    \begin{agdalisting*}
      \ExecuteMetaData[agda/Cardinality/Finite/SplitEnumerable.tex]{is-split-inj}
    \end{agdalisting*}
  \end{minipage}
  \begin{minipage}[t]{.19\textwidth} \setstretch{1.1}

      Def.~\ref{split-enum-def} (\AgdaDatatype{\ensuremath{\mathcal{E}!}}) \\
      Eqn.~\ref{container-membership} (\AgdaDatatype{\ensuremath{\in}}) \\
      Eqn.~\ref{sp-surj-eqn}  \\
      Eqn.~\ref{list-def} (\(\AgdaDatatype{List}\)) \\
      Eqn.~\ref{container-interp}  \\
      Reassociation \\
      Eqn.~\ref{sp-surj-arrow-eqn}
  \end{minipage}
\end{proof}
In the above proof syntax the
\(\AgdaDatatype{\ensuremath{\equiv \langle{} \rangle{} }}\) connects lines which
are definitionally equal, i.e. they are ``obviously'' equal from the type
checker's perspective.
Clearly, only one line isn't a definitional equality: 
\begin{agdalisting}
  \ExecuteMetaData[agda/Data/Sigma/Properties.tex]{reassoc}
\end{agdalisting}
This means that we could have in fact written the whole proof as follows:
\begin{agdalisting}
  \ExecuteMetaData[agda/Cardinality/Finite/SplitEnumerable.tex]{split-enum-is-split-surj-short}
\end{agdalisting}
The simplicity of this proof, by the way, is why we preferred the
container-based definition of lists over the traditional one.
\subsection{Instances} \todo{More intro to this}
We will now provide some example proofs of types which are split enumerable.
\begin{lemma}
  \(\bot\), \(\top\), and \(\AgdaDatatype{Bool}\) are split enumerable.
\end{lemma}
\begin{proof}
  These three types are quite obviously finite: we will show only the proof of
  finiteness for \(\AgdaDatatype{Bool}\) here for brevity's sake.
  \begin{agdalisting} \label{bool-is-split-enumerable}
    \ExecuteMetaData[agda/Cardinality/Finite/SplitEnumerable.tex]{bool-inst}
  \end{agdalisting}
\end{proof}

As a slightly more complex example, consider the \(\AgdaDatatype{Fin}\) type
we've been using.
Remember that split enumerability is in fact the same as a split surjection from
\(\AgdaDatatype{Fin}\) (Lemma~\ref{split-enum-is-split-surj}): to show that
\(\AgdaDatatype{Fin}\) is split enumerable, then, we need only show that it has
a split surjection from itself.
We'll prove the following slightly more general statement:
\begin{lemma}
  Every type has a split surjection into itself: the identity function.
\end{lemma}
\begin{proof}
  We provide the simple proof here in Agda:
  \begin{agdalisting}
    \ExecuteMetaData[agda/Function/Surjective/Properties.tex]{split-surj-ident}
  \end{agdalisting}
\end{proof}
\subsection{Decidable Equality}
One thing that characterises all split enumerable types is that they are all
\emph{discrete}.
We will see later that this has implications for the space of types we're
dealing with, but for now it simply provides a useful function on split
enumerable types.
\begin{definition}[Discrete Types]
  A discrete type is one with decidable equality.
  \begin{agdalisting}
    \ExecuteMetaData[agda/Relation/Nullary/Discrete/Base.tex]{discrete-def}
  \end{agdalisting}
\end{definition}
Of course we also will need to explain what we mean by ``decidable''.
\begin{definition}[Decidable Types]
  If some proposition is decidable it means that it is proven true or false.
  We can encapsulate this with a data type\footnotemark:
  \begin{agdalisting}
    \ExecuteMetaData[agda/Snippets/Dec.tex]{dec-def}
  \end{agdalisting}
  A proof of type \(\AgdaDatatype{Dec}\;A\) is a proof that \(A\) is either true
  or false.
\end{definition}
\footnotetext{This is not, as it happens, the definition of
  \(\AgdaDatatype{Dec}\) you will find in Agda's standard library.
The version of the type that exists there is a good deal more complicated (for
computational reasons),
although its interface behaves}

So now we have our task in focus: we need to prove that split enumerability
implies decidable equality.
We will do that by first describing split enumerability in terms of injections.
\todo{This proof could be presented a lot better, I think}
\begin{definition}[Injections]
  Injective functions are more straightforward to define constructively than
  surjective ones:
  \begin{agdalisting}
    \ExecuteMetaData[agda/Function/Injective/Base.tex]{injective}
  \end{agdalisting} \vspace{-1.5\baselineskip}
  \begin{agdalisting}
    \ExecuteMetaData[agda/Function/Injective/Base.tex]{injection}
  \end{agdalisting}
\end{definition}
\begin{lemma} \label{split-surj-to-inj}
  A split-surjection from \(A\) to \(B\) implies an injection from \(B\) to
  \(A\).
  \begin{agdalisting}
    \ExecuteMetaData[agda/Function/Surjective/Properties.tex]{surj-to-inj}
  \end{agdalisting}
\end{lemma}
\begin{proof} \let\qed\relax

  \begin{minipage}[t]{.9\textwidth}\vspace{-\baselineskip}
    \begin{agdalisting*}
      \ExecuteMetaData[agda/Function/Surjective/Properties.tex]{surj-to-inj-proof}
    \end{agdalisting*}
  \end{minipage}
\end{proof}
\begin{lemma} \label{inj-discrete}
  For any type \(A\) which injects into a discrete type \(B\), \(A\) is
  discrete.
  \begin{agdalisting}
    \ExecuteMetaData[agda/Function/Injective/Properties.tex]{inj-discrete}
  \end{agdalisting}
\end{lemma}

\begin{lemma} \label{discrete-surj}
  If a discrete type \(A\) has a split surjection into some type \(B\), \(B\) is
  discrete.
  \begin{agdalisting}
    \ExecuteMetaData[agda/Function/Surjective/Properties.tex]{discrete-surj}
  \end{agdalisting}
\end{lemma}
\begin{proof}
  This proof is can be straightforwardly derived from lemmas
  \ref{split-surj-to-inj} and \ref{inj-discrete}.
\end{proof}

\begin{lemma} \label{split-enum-discrete}
  Every split enumerable type is discrete.
\end{lemma}
\begin{proof}
  Let \(A\) be a split enumerable type.
  By lemma~\ref{split-enum-is-split-surj}, there is a surjection from
  \(\AgdaDatatype{Fin}\;n\) for some \(n\).
  Also, we know that \(\AgdaDatatype{Fin}\;n\) is discrete (proven in our
  formalisation).
  Therefore, by lemma~\ref{discrete-surj}, \(A\) is discrete.
\end{proof}
\section{Manifest Bishop Finiteness} \label{manifest-bishop-finiteness}
We mentioned in the introduction that occasionally in constructive mathematics
proofs will contain ``too much'' information.
With split enumerability we can see an instance of this.
Consider the following proof of the finiteness of bool:
\begin{agdalisting} \label{op-slop}
  \ExecuteMetaData[agda/Cardinality/Finite/SplitEnumerable.tex]{bool-slop}
\end{agdalisting}
There is an extra \AgdaInductiveConstructor{false} at the end of the support
list.
There's nothing terribly wrong with that: it is still a valid proof of
finiteness, after all, but it does mean that this proof has some extra
information which we didn't necessarily intend to encode.

>>>>

One of the nice things about containers is it gives us a generic way to define
``membership'':
\begin{multicols}{2}
  \begin{agdalisting} \label{container-membership}
    \ExecuteMetaData[agda/Container/Membership.tex]{membership-def}
  \end{agdalisting} \columnbreak
  \begin{agdalisting}
    \ExecuteMetaData[agda/Snippets/Introduction.tex]{fiber}
  \end{agdalisting}
\end{multicols} \vspace{-1\baselineskip}\noindent 
Here we're using the homotopy-theory notion of a \AgdaDatatype{fiber} to define
membership: a fiber for some function \(f\) and some point \(y\) in its codomain
is a value \(x\) and a proof that \(f \; x \;\AgdaDatatype{≡}\;y\).
Membership also makes more sense when described concretely in terms of lists:
\(x\;\AgdaDatatype{\ensuremath{\in}}\;\mathit{xs}\) means ``there is an index
into \(\mathit{xs}\) such that the index points at an item equal to \(x\)''.
\subsection{Split Surjections}
Now that we have our terms defined, let's look a little at how split
enumerability relates to more traditional, classical notions of finiteness.
In a classical setting we likely wouldn't mention ``lists'' or the like, and
would instead define finiteness based on the existence of some injection or
surjection, say a surjection from a finite prefix of the natural numbers.
In HoTT, surjections (or, more precisely, \emph{split} surjections
\cite[definition 4.6.1]{hottbook}), are defined like so:
\begin{multicols}{2}
  \begin{agdalisting} \label{sp-surj-eqn}
    \ExecuteMetaData[agda/Function/Surjective/Base.tex]{split-surjective}
  \end{agdalisting} \columnbreak
  \begin{agdalisting} \label{sp-surj-arrow-eqn}
    \ExecuteMetaData[agda/Function/Surjective/Base.tex]{split-surjection}
  \end{agdalisting} 
\end{multicols}\vspace{-1\baselineskip}
As it turns out, our definition of finiteness here is precisely the same as a
surjection-based one, in quite a deep way!
\begin{lemma} \label{split-enum-is-split-surj}%
  A proof of split enumerability is equivalent to a split surjection from a
  finite prefix of the natural numbers.
  \begin{agdalisting}
    \ExecuteMetaData[agda/Cardinality/Finite/SplitEnumerable.tex]{is-split-inj-type}
  \end{agdalisting}
\end{lemma}
\begin{proof} \let\qed\relax \vspace{-1\baselineskip}
  \begin{minipage}[t]{.7\textwidth}\vspace{-\baselineskip}
    \begin{agdalisting*}
      \ExecuteMetaData[agda/Cardinality/Finite/SplitEnumerable.tex]{is-split-inj}
    \end{agdalisting*}
  \end{minipage}
  \begin{minipage}[t]{.19\textwidth} \setstretch{1.1}

      Def.~\ref{split-enum-def} (\AgdaDatatype{\ensuremath{\mathcal{E}!}}) \\
      Eqn.~\ref{container-membership} (\AgdaDatatype{\ensuremath{\in}}) \\
      Eqn.~\ref{sp-surj-eqn}  \\
      Eqn.~\ref{list-def} (\(\AgdaDatatype{List}\)) \\
      Eqn.~\ref{container-interp}  \\
      Reassociation \\
      Eqn.~\ref{sp-surj-arrow-eqn}
  \end{minipage}
\end{proof}
In the above proof syntax the
\(\AgdaDatatype{\ensuremath{\equiv \langle{} \rangle{} }}\) connects lines which
are definitionally equal, i.e. they are ``obviously'' equal from the type
checker's perspective.
Clearly, only one line isn't a definitional equality: 
\begin{agdalisting}
  \ExecuteMetaData[agda/Data/Sigma/Properties.tex]{reassoc}
\end{agdalisting}
This means that we could have in fact written the whole proof as follows:
\begin{agdalisting}
  \ExecuteMetaData[agda/Cardinality/Finite/SplitEnumerable.tex]{split-enum-is-split-surj-short}
\end{agdalisting}
The simplicity of this proof, by the way, is why we preferred the
container-based definition of lists over the traditional one.
\subsection{Instances}
To actually show that a type \(A\) is finite amounts to constructing a term of
type \(\AgdaDatatype{\ensuremath{\mathcal{E}!}}\;A\).
For simple types like \AgdaDatatype{Bool}, that is simple:
\begin{agdalisting} \label{bool-is-split-enumerable}
  \ExecuteMetaData[agda/Cardinality/Finite/SplitEnumerable.tex]{bool-inst}
\end{agdalisting}

As a slightly more complex example, consider the \(\AgdaDatatype{Fin}\) type
we've been using.
Remember that split enumerability is in fact the same as a split surjection from
\(\AgdaDatatype{Fin}\) (Lemma~\ref{split-enum-is-split-surj}): to show that
\(\AgdaDatatype{Fin}\) is split enumerable, then, we need only show that it has
a split surjection from itself.
We'll prove the following slightly more general statement:
\begin{agdalisting}
  \ExecuteMetaData[agda/Function/Surjective/Properties.tex]{split-surj-ident}
\end{agdalisting}
\subsection{Decidable Equality}
One thing that characterises all split enumerable types is that they are all
\emph{discrete}, i.e. they have decidable equality.  \todo{All multicol agda code should probably be vertically centred like this}
\begin{multicols}{2}  \null \vfill
  \begin{agdalisting}
    \ExecuteMetaData[agda/Relation/Nullary/Discrete/Base.tex]{discrete-def}
  \end{agdalisting} \vfill \null \columnbreak
  \begin{agdalisting}
    \ExecuteMetaData[agda/Snippets/Dec.tex]{dec-def}
  \end{agdalisting}
\end{multicols} \vspace{-\baselineskip} \noindent
We will see later that this has implications for the space of types we're
dealing with, but for now it simply provides a useful function on split
enumerable types.

To prove that split enumerability implies decidable equality we'll take a quick
detour through injections.
\begin{multicols}{2}
  \begin{agdalisting}
    \ExecuteMetaData[agda/Function/Injective/Base.tex]{injective}
  \end{agdalisting} \columnbreak
  \begin{agdalisting}
    \ExecuteMetaData[agda/Function/Injective/Base.tex]{injection}
  \end{agdalisting}
\end{multicols} \vspace{-\baselineskip} \noindent
These are useful because we know that any type which injects into a discrete
type is itself discrete:
\begin{agdalisting}
  \ExecuteMetaData[agda/Function/Injective/Properties.tex]{inj-discrete}
\end{agdalisting}
And we can turn a split surjection from \(A\) to \(B\) into an injection from
\(B\) to \(A\): \todo{make these side-wrapped floats?}
\begin{agdalisting}
  \ExecuteMetaData[agda/Function/Surjective/Properties.tex]{surj-to-inj}
\end{agdalisting}
Yielding a simple proof that any type with a split surjection from a discrete
type is itself discrete:
\begin{agdalisting}
  \ExecuteMetaData[agda/Function/Surjective/Properties.tex]{discrete-surj}
\end{agdalisting}
Since split enumerability is really just a split surjection from
\AgdaDatatype{Fin}, and since we know that \AgdaDatatype{Fin} is discrete, the
overall proof resolves quite simply:
\begin{agdalisting} \label{split-enum-discrete}
  \ExecuteMetaData[agda/Cardinality/Finite/SplitEnumerable.tex]{split-is-discrete}
\end{agdalisting}
\section{Manifest Bishop Finiteness} \label{manifest-bishop-finiteness}
We mentioned in the introduction that occasionally in constructive mathematics
proofs will contain ``too much'' information.
With split enumerability we can see an instance of this.
Consider the following proof of the finiteness of bool:
\begin{agdalisting} \label{op-slop}
  \ExecuteMetaData[agda/Cardinality/Finite/SplitEnumerable.tex]{bool-slop}
\end{agdalisting}
There is an extra \AgdaInductiveConstructor{false} at the end of the support
list.
There's nothing terribly wrong with that: it is still a valid proof of
finiteness, after all, but it does mean that this proof has some extra
information which we didn't necessarily intend to encode.
